``Interestingness'' in Mathematical Theory Exploration

### 2019-11-24 PhDThesis/litreview_interestingness.tex
Some rearranging and expansion

When we \emph{generalise} a statement $S$, we obtain a new statement $S'$ of
which $S$ is a special case. Although it seems counterintuitive, a generalised
statement can sometimes be \emph{easier} to prove than the original. This arises
often in inductive proofs (e.g. when reasoning about tail-recursive
definitions~\cite{kapur2003automatic}), since the specific obligations which
arise in the proof may be incompatible with the available inductive hypotheses.
An informative example is given by Boyer and Moore of the associativity of
multiplication in ACL2~\cite{boyer1983proof}:

@@

Generality hence contributes to a conjecture's \emph{utility}, and also (by
definition) to its \emph{applicability}. It is still important to balance
against other concerns, since blindly generalising \emph{all} statements of
proof obligations we come across would result in conjectures so strong that they
are unprovable, or even false; which would make them much less interesting.

@@

The interestingness of a statement can also be characterised by \emph{analogy}
to existing interesting statements. By finding lemmas analogous to those of a
different theory, we may be able to re-use proof tactics and other forms of
meta-programming across both.

Existing theory exploration systems have been successfully applied to this
problem, however the use of pure exploration misses opportunities to
\emph{focus} the search, since we know which lemmas are used in those theories
where a technique succeeded. If we can find an analogy to map from such solved
problems to a currently-unproven goal, we can infer the approximate form of the
lemmas we require, and target these specifically.

@@

The interestingness of such generated lemmas derives from both the existence of
the original lemma (which is presumably interesting enough to prove and
subsequently explore), that of the target theorem it may help to prove, and the
novelty provided by reframing existing concepts in a new domain. The running
examples for demonstrating \textsc{ACL2(ml)} are equivalence theorems for
tail-recursive and non-tail-recursive calculations\footnote{A tail-recursive
  function can be executed in constant space using a loop, whereas recursion in
  non-tail positions may require a growing number of stack frames or nested
  closures.}), as well as the effect of repeating certain list operations:

@@

% TODO Include something about intrinsic motivation, compression progress,
% variance among experts (decision tree stuff as well as more established
% splitting functions stuff)

>>>>

When we \emph{generalise} a statement $S$, we obtain a new statement $S'$ of
which $S$ is a special case. Although it seems counterintuitive, a generalised
statement can sometimes be \emph{easier} to prove than the original. This arises
often in inductive proofs (e.g. when reasoning about tail-recursive
definitions~\cite{kapur2003automatic}\footnote{Tail-recursive functions are
  desirable due to their constant space usage, unlike recursion in non-tail
  positions which requires a growing number of stack frames or nested closures.
  Reasoning about non-tail recursive definitions is often simpler, since they
  don't expose accumulator arguments, so equivalence ), since the specific obligations which
arise in the proof may be incompatible with the available inductive hypotheses.
An informative example is given by Boyer and Moore of the associativity of
multiplication in ACL2~\cite{boyer1983proof}:

@@

Generality hence contributes to a conjecture's \emph{utility}, and also (by
definition) to its \emph{applicability}. It is still important to balance
against other concerns, since blindly generalising \emph{all} statements or
proof obligations we come across would result in conjectures so strong that they
are unprovable, or even false; which would make them much less interesting.

@@

The interestingness of a statement can also be characterised by \emph{analogy}
to existing interesting statements. By finding lemmas analogous to those of a
different theory, we may be able to re-use proof tactics and other forms of
meta-programming across both. In this case an analogy would map from a solved
problem to a currently-unproven goal, and applying this map to the lemmas
appearing in the existing proof allows us to infer the approximate form of the
lemmas required for the new theorem.

@@

The interestingness of such generated lemmas derives from both the existence of
the original lemma (which is presumably interesting enough to prove and
subsequently explore), that of the target theorem it may help to prove, and the
novelty provided by reframing existing concepts in a new domain. The running
examples for demonstrating \textsc{ACL2(ml)} are equivalence theorems for
tail-recursive and their non-tail-recursive counterparts, as well as the effect
of repeating certain list operations:

@@

\subsection{Tests}

Since we are working in the domain of Haskell programs, an abundant source of
statements are available in the form of \emph{tests}. For manually-written
tests, the effort required to write them implies that they must be of some
interest to their author. Many forms of test are \emph{not} suitable for our
purposes, such as \emph{unit tests} which are always trivially provable by
$\beta$-reduction; or \emph{integration tests}, which depend on the behaviour of
the external environment. One reason to study Haskell is its widespread use of
\emph{property checking}, which does give us useful data in the form of
universally quantified statements. Many Haskell property checkers exist, based
on random testing (\quickcheck{}~\cite{claessen2011quickcheck} and
\smartcheck{}~\cite{pike2014smartcheck}), enumeration
(\smallcheck{}~\cite{runciman2008smallcheck}), observation
(\lazysmallcheck{}~\cite{reich2013advances}) and logic programming
(\sparsecheck{}~\cite{sparsecheck}).

Thankfully the major differences between these systems are in the way they
instantiate test arguments; their representations of properties are largely the
same (modulo renaming of functions and types): as Haskell functions with free
variables encoded as function arguments.

\subsection{Artificial Curiosity} \label{sec:intrinsic}

\begin{figure}
  \centering
  \begin{tikzpicture}[scale=0.75]
      % The image, for reference
      % \node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=\textwidth]{wundt.png}};

      % Axes
      \draw[black,ultra thick,->] (1,0.8) -- (1,  9)   node[midway, above, sloped] {HEDONIC VALUE};     % y axis
      \draw[black,ultra thick,->] (1,4)   -- (11, 4);                                                   % x axis
      \path                       (1,0.8) -- (11, 0.8) node[midway, below]         {AROUSAL POTENTIAL}; % x axis label

      % Curve. The numbers come from tracing over wundt.png
      \draw[black,ultra thick] (3, 4)
           to[out=0,   in=240] (5.05, 6)
           to[out=60,  in=180] (6.5,  8.2)
           to[out=0,   in=105] (8.1,  6)
           to[out=-75, in=120] (8.8,  4)
           to[out=-60, in=150] (11.4, 0.8);

      % This version is closer, but a little jagged
      \iffalse
      \draw[black,ultra thick] (3, 4)
           to[out=0,   in=240] (5.05, 6)
           to[out=60,  in=225] (6,    8)
           to[out=45,  in=180] (6.5,  8.2)
           to[out=0,   in=135] (7.2,  8)
           to[out=-45, in=105] (8.1,  6)
           to[out=-75, in=120] (8.8,  4)
           to[out=-60, in=150] (11.4, 0.8);
      \fi
  \end{tikzpicture}

  \caption{The Wundt curve, reproduced from~\cite{berlyne1970novelty}. The axes
    ``hedonic value'' and ``arousal potential'' are described as covering
    \textquote{reward value\dots preference or pleasure}, and \textquote{all the
      stimulus properties that tend to raise arousal, including novelty and
      complexity}, respectively.}
  \label{fig:wundt}
\end{figure}

Another perspective on what is ``interesting'' comes from the artificial
intelligence tasks of optimisation and reinforcement learning. Common pitfalls
for such systems are highly non-convex problems (trapping optimisers in local
minima), and sparse information (such as flat gradients; or highly delayed
rewards, like a single win/lose reward at the end of a game). To mitigate these problems, various forms of internally-generated (or \emph{intrinsic}) rewards
can be incorporated into a system. Such schemes provide reward for interesting
discoveries, which encourages exploration in lieu of external direction, and are
known as \emph{Artificial Curiosity}~\cite{schmidhuber2006developmental} (AC).

The unifying principle of AC methods is to force systems away from data which
are not amenable to learning; either because they are so familiar that there is
nothing left to learn, or so unfamiliar that they are unintelligible. The
resulting behaviour is characterised by the \emph{Wundt curve} (shown in Figure \ref{fig:wundt}) \footnote{In practice, many measures
  avoid negative values for simplicity, in which cases we replace all negative
  points on the curve with zero.}, which has been used in psychology to explain
human aesthetics and preferences \cite{berlyne1970novelty}. This same behaviour
may be applicable to the theorems produced by a theory exploration system.

We can divide AC approaches into two groups: those which make \emph{explicit}
use of interestingness, learning from signals which follow a Wundt curve; whilst
\emph{implicit} approaches modify the \emph{output} of their learning
algorithm(s), to engineer the overall system behaviour to follow a Wundt curve
as an emergent property.

A framework encompassing many examples of the explicit approach is given
in~\cite{oudeyer2007intrinsic} in the context of reinforcement learning; for
comparison, many similar measures are surveyed in a data mining context
in~\cite{geng2006interestingness}. Many more reinforcement learning examples can
be found in~\cite{Kaplan2006, Lipson2007, Luciw2011, Macedo2000,
  Ramik.Sabourin.Madani:2013, Roa.Kruijff.Jacobsson:2009, Schmidhuber:1991,
  oudeyer2004intelligent}; whilst more general descriptions are given
in~\cite{Schaul.Sun.Wierstra.ea:2011, Scott1989, maher2008achieving}, which may
be more amenable to Theory Exploration setting.

Many of these reward signals are based on information theory, with a prominent
example being \emph{compression progress}: given a compressed representation of
our previous observations, the ``progress'' is the space saved if we include the
current observation. Observations which are incompressible or trivially
compressible don't save any space, whilst observations which provide new
insights into the structure of past experience can provide a space saving when
compressed together. This seems particularly relevant for identifying
interesting theorems: those new theorems (``observations'') which shorten the
proofs of previously discovered theorems may be more general, more powerful and
therefore more \emph{interesting}. In fact this is very similar to
\quickspec{}'s interestingness criterion.

Another example of explicit Artificial Curiosity is given
in~\cite{Hester.Stone:2012}, where world states which cause \emph{disagreement}
among a population of decision trees (a \emph{random
  forest}~\cite{randomforests}) are considered interesting. Since the models
make stochastic predictions, the disagreement follows a Wundt curve as the
complexity of state transitions increases: for parts of the state space which
have been fully learned, the models will agree on accurate predictions; for
parts which are unlearnable, the models cannot infer any structure, and will
converge to reporting the same average value. Whilst the latter predictions may
not be \emph{accurate}, they will be \emph{in agreement}, hence pushing down the
interestingness of states which are too complex.

Many examples of the implicit case are based on \emph{coevolution}: rewarding
one part of the system for exploiting another part, and vice versa. In~\cite{Schmidhuber1999} a pair of learning algorithms place virtual ``bets''
on the outcome of actions, and the winner is rewarded at the expense of the
loser. Due to the risk involved, each algorithm will only bet when it is
confident in its prediction, and bets will only be actioned when each algorithm
is confident in a \emph{different} outcome. The overall behaviour of this system
is therefore similar to the explicit measure of disagreement used in the random
forest example.

More recent work on \emph{Generative Adversarial Networks} (GANs) trains a
``discriminator'' program to distinguish real training examples (usually images)
from synthetic data made by a ``generator'' program. Both programs are
differentiable (usually some type of Artificial Neural Network), so the coupling
between the generator's output and the discriminator's input can be used not
only to feed synthetic data ``forwards'', but also to back-propagate the
discriminator's errors (which it is trying to minimise) into the generator
(which then tries to maximise, exploiting the discriminator by synthesising more
realistic data).

Another example of implicit AC is the ``darwinian brain'' of Fernando et
al.~\cite{fernando2013design1, fernando2013design2}. This coevolves a
population of problem generators and problem solvers, rewarding the solvers
based on their speed, and rewarding the generators based on the \emph{variance}
of the solvers' speed. This avoids trivial problems (which all solvers can
quickly overcome) and complex problems (which no solver can manage), and focuses
on those with the most possibility for learning. The role of problem generator
is similar to that of Theory Exploration tools, with solvers being Automated
Theorem Provers.

% TODO Include something about
% variance among experts (decision tree stuff as well as more established
% splitting functions stuff)

% TODO: Also mention the intrinsic reward idea about easily accessible regions
% (e.g. balancing a pole because it's easier to get from there to anywhere else)
